---
title: "MEDIPS workflow"
author: "Alvaro Bautista Abad de la Cruz"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: cerulean
    highlight: tango
  word_document:
    toc: true
  pdf_document:
    toc: true
    toc_depth: '4'
---

# Raw and processed MeDIP-seq data are available from the NCBI Gene Expression Omnibus under Series accession GSE338843.


```{r portable-paths, include=FALSE}
.current_input <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
if (!is.null(.current_input) && nzchar(.current_input)) {
  SCRIPT_DIR <- dirname(normalizePath(.current_input, winslash = "/", mustWork = FALSE))
} else {
  SCRIPT_DIR <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
}
BUNDLE_ROOT <- normalizePath(file.path(SCRIPT_DIR, "..", ".."), winslash = "/", mustWork = FALSE)
DATA_ROOT <- file.path(BUNDLE_ROOT, "data")
OUTPUT_ROOT <- file.path(BUNDLE_ROOT, "generated_outputs")
ANALYSIS_NAME <- basename(SCRIPT_DIR)
DATA_DIR <- file.path(DATA_ROOT, ANALYSIS_NAME)
OUTPUT_DIR <- file.path(OUTPUT_ROOT, ANALYSIS_NAME)
TABLE_DIR <- file.path(DATA_DIR, "workspace_and_tables")
dir.create(OUTPUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
input_file <- function(...) file.path(DATA_DIR, ...)
output_file <- function(...) file.path(OUTPUT_DIR, ...)
table_file <- function(...) file.path(TABLE_DIR, ...)
input_or_output_file <- function(...) {
  output_path <- output_file(...)
  if (file.exists(output_path)) output_path else input_file(...)
}
knitr::opts_knit$set(root.dir = OUTPUT_DIR)
RUN_STARTED_AT <- Sys.time()
```
```{r setup, include=FALSE}
knitr_cache_path <- paste0(file.path(OUTPUT_DIR, "knitr_cache_ws400"), "/")
knitr_fig_path <- paste0(file.path(OUTPUT_DIR, "knitr_figures_ws400"), "/")
dir.create(knitr_cache_path, showWarnings = FALSE, recursive = TRUE)
dir.create(knitr_fig_path, showWarnings = FALSE, recursive = TRUE)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  autodep = TRUE,
  cache.lazy = FALSE,
  cache.path = knitr_cache_path,
  fig.path = knitr_fig_path
)

options(stringsAsFactors = FALSE)
options(timeout = 600)

# This notebook exports the analysis tables supporting the manuscript figures.

```

## 1. Required packages

```{r check-packages}
required_pkgs <- c(
  "BSgenome", "MEDIPS", "BSgenome.Rnorvegicus.UCSC.rn6", "GenomicRanges",
  "Rsamtools", "rtracklayer", "AnnotationDbi", "S4Vectors", "BiocGenerics",
  "GenomeInfoDb", "magrittr", "dplyr", "readxl", "writexl",
  "clusterProfiler", "org.Rn.eg.db", "stringr", "tidyr", "purrr", "tibble"
)

missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]

if (length(missing_pkgs) > 0) {
  stop(
    "Missing packages before knitting:\n",
    paste(missing_pkgs, collapse = ", ")
  )
}
```

# 2. Initial setup

```{r, results= 'hide'}
library("BSgenome") # Load BSgenome
library(MEDIPS) # Load MEDIPS
library(BSgenome.Rnorvegicus.UCSC.rn6) # Load the rat reference genome (rn6)
library(Rsamtools) # Load Rsamtools for SAM/BAM processing
```

The following objects store the paths to the BAM files.

> **Note — BAM files are not distributed with this archive.**
> The coordinate-sorted rn6 BAM/BAI files (~32 GB) are omitted from the Zenodo
> record because of their size. Obtain them before running this chunk by either
> (1) downloading the MeDIP-seq FASTQ files from GEO (accession GSE-XXXXXX) and
> aligning them to UCSC rn6 with HISAT2 v2.1.0 (see the README, "Reproducing the
> alignment"), or (2) requesting the aligned BAM files from the corresponding
> author. Place the resulting `.bam`/`.bai` files, keeping the file names used
> below, under `data/MeDIP/Aligned_Results_BAM/`.

```{r}
# -----------------------------------------------------------------------------
# BAM paths
# -----------------------------------------------------------------------------
bam_dir <- input_file("Aligned_Results_BAM")

if (!dir.exists(bam_dir)) {
  stop(
    "BAM directory not found: ", bam_dir, "\n",
    "The coordinate-sorted rn6 BAM files are NOT included in the Zenodo archive ",
    "(~32 GB). Obtain them in one of two ways:\n",
    "  1. Download the MeDIP-seq FASTQ from GEO (accession GSE-XXXXXX) and align ",
    "to UCSC rn6 with HISAT2 v2.1.0 (see README, 'Reproducing the alignment'), ",
    "then place the sorted .bam/.bai files in this folder; or\n",
    "  2. Request the aligned BAM files from the corresponding author."
  )
}
```

### Female control

```{r, results= 'hide'}
Female.control.1.bam.file.MeDIP <- file.path(bam_dir, "Female-control-1-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Female.control.2.bam.file.MeDIP <- file.path(bam_dir, "Female-control-2-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Female.control.3.bam.file.MeDIP <- file.path(bam_dir, "Female-control-3-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Female.control.bam.file.Input   <- file.path(bam_dir, "Female-control-input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
```

### Female CSF1

```{r,  results= 'hide'}
Female.CSF1.1.bam.file.MeDIP <- file.path(bam_dir, "Female-CSF-1-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Female.CSF1.2.bam.file.MeDIP <- file.path(bam_dir, "Female-CSF-2-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Female.CSF1.3.bam.file.MeDIP <- file.path(bam_dir, "Female-CSF-3-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Female.CSF1.bam.file.Input   <- file.path(bam_dir, "Female-CSF-input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
```

### Male control

```{r,  results= 'hide'}
Male.control.1.bam.file.MeDIP <- file.path(bam_dir, "Male-control-1-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Male.control.2.bam.file.MeDIP <- file.path(bam_dir, "Male-control-2-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Male.control.3.bam.file.MeDIP <- file.path(bam_dir, "Male-control-3-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Male.control.bam.file.Input   <- file.path(bam_dir, "Male-control-input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
```

### Male CSF1

```{r,  results= 'hide'}
Male.CSF1.1.bam.file.MeDIP <- file.path(bam_dir, "Male-CSF-1-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Male.CSF1.2.bam.file.MeDIP <- file.path(bam_dir, "Male-CSF-2-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Male.CSF1.3.bam.file.MeDIP <- file.path(bam_dir, "Male-CSF-3-IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
Male.CSF1.bam.file.Input   <- file.path(bam_dir, "Male-CSF-input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam")
```

```{r}
bam_files <- c(
  Female.control.1.bam.file.MeDIP,
  Female.control.2.bam.file.MeDIP,
  Female.control.3.bam.file.MeDIP,
  Female.control.bam.file.Input,
  Female.CSF1.1.bam.file.MeDIP,
  Female.CSF1.2.bam.file.MeDIP,
  Female.CSF1.3.bam.file.MeDIP,
  Female.CSF1.bam.file.Input,
  Male.control.1.bam.file.MeDIP,
  Male.control.2.bam.file.MeDIP,
  Male.control.3.bam.file.MeDIP,
  Male.control.bam.file.Input,
  Male.CSF1.1.bam.file.MeDIP,
  Male.CSF1.2.bam.file.MeDIP,
  Male.CSF1.3.bam.file.MeDIP,
  Male.CSF1.bam.file.Input
)

missing_bam_files <- bam_files[!file.exists(bam_files)]

if (length(missing_bam_files) > 0) {
  stop(
    "The following BAM files are missing:\n",
    paste(missing_bam_files, collapse = "\n")
  )
} else {
  message("All BAM files were located successfully.")
}
```


```{r,  results= 'hide'}
BSgenome="BSgenome.Rnorvegicus.UCSC.rn6"
extend=300
shift=0
ws=400
uniq= 1e-3
chr.select=c("chr1","chr2","chr3","chr4","chr5","chr6","chr7","chr8","chr9","chr10","chr11","chr12","chr13","chr14","chr15","chr16","chr17","chr18","chr19","chr20","chrM","chrY","chrX")

wd <- OUTPUT_DIR # Portable working/output directory
```


# 4. Analysis

### 4.1 Data import and preprocessing

#### MEDIPS

##### Female control

```{r, results= 'hide'}
# Each condition is assembled from individual animals; samples remain biological replicates.
Female_Control_MeDIP = MEDIPS.createSet(file = Female.control.1.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select)

Female_Control_MeDIP = c(Female_Control_MeDIP, MEDIPS.createSet(file = Female.control.2.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select))

Female_Control_MeDIP = c(Female_Control_MeDIP, MEDIPS.createSet(file = Female.control.3.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select))
```

##### Female CSF

```{r, results= 'hide'}
Female_CSF_MeDIP = MEDIPS.createSet(file = Female.CSF1.1.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select)

Female_CSF_MeDIP = c(Female_CSF_MeDIP, MEDIPS.createSet(file = Female.CSF1.2.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select))

Female_CSF_MeDIP = c(Female_CSF_MeDIP, MEDIPS.createSet(file = Female.CSF1.3.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select))
```

##### Male control

```{r,  results= 'hide'}
Male_Control_MeDIP = MEDIPS.createSet(file = Male.control.1.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select)

Male_Control_MeDIP = c(Male_Control_MeDIP, MEDIPS.createSet(file = Male.control.2.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select))

Male_Control_MeDIP = c(Male_Control_MeDIP, MEDIPS.createSet(file = Male.control.3.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select))
```

##### Male CSF

```{r,  results= 'hide'}
Male_CSF_MeDIP = MEDIPS.createSet(file = Male.CSF1.1.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select)

Male_CSF_MeDIP = c(Male_CSF_MeDIP, MEDIPS.createSet(file = Male.CSF1.2.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select))

Male_CSF_MeDIP = c(Male_CSF_MeDIP, MEDIPS.createSet(file = Male.CSF1.3.bam.file.MeDIP, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select))
```

#### INPUTS

```{r,  results= 'hide'}
# Input DNA is used as the control set for MeDIP normalization.
Female_control_Input = MEDIPS.createSet(file = Female.control.bam.file.Input, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select)

Female_CSF_Input = MEDIPS.createSet(file = Female.CSF1.bam.file.Input, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select)

Male_control_Input = MEDIPS.createSet(file = Male.control.bam.file.Input, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select)

Male_CSF_Input = MEDIPS.createSet(file = Male.CSF1.bam.file.Input, BSgenome = BSgenome, extend = extend, shift = shift, uniq = uniq, window_size = ws, chr.select = chr.select)
```

#### Normalization (CG dinucleotide content)

```{r,  results= 'hide'}
CS_Male = MEDIPS.couplingVector(pattern = "CG", refObj = Male_Control_MeDIP) # Reference control object for CG-content calculation.
```

```{r,  results= 'hide'}
# The coupling vector depends only on genome coordinates and window size.
# All MEDIPS sets use rn6, the same chromosomes, and 400-bp windows.
CS_Female = CS_Male
```

```{r,  results= 'hide'}
CS_Male_CSF = CS_Male
```

### 4.2 Differential coverage and methylation profiles

#### Female CSF1 vs Female control

```{r,  results= 'hide'}
Female_csf_vs_Female_control_mr.edgeR = MEDIPS.meth(MSet1 = Female_CSF_MeDIP, MSet2 = Female_Control_MeDIP, CSet = CS_Female, ISet1 = Female_CSF_Input, ISet2 = Female_control_Input, p.adj = "none", diff.method = "edgeR", MeDIP = T, CNV = T, minRowSum = 10)
```

#### Male CSF1 vs Male control

```{r,  results= 'hide'}
Male_csf_vs_Male_control_mr.edgeR = MEDIPS.meth(MSet1 = Male_CSF_MeDIP, MSet2 = Male_Control_MeDIP, CSet = CS_Male, ISet1 = Male_CSF_Input, ISet2 = Male_control_Input, p.adj = "none", diff.method = "edgeR", MeDIP = T, CNV = T, minRowSum = 10)
```

#### Female control vs Male control

```{r,  results= 'hide'}
Female_control_vs_Male_control_mr.edgeR = MEDIPS.meth(MSet1 = Female_Control_MeDIP, MSet2 = Male_Control_MeDIP, CSet = CS_Male, ISet1 = Female_control_Input, ISet2 = Male_control_Input, p.adj = "none", diff.method = "edgeR", MeDIP = T, CNV = T, minRowSum = 10)
```

#### Female CSF vs Male CSF

```{r,  results= 'hide'}
Female_csf_vs_Male_csf_mr.edgeR = MEDIPS.meth(MSet1 = Female_CSF_MeDIP, MSet2 = Male_CSF_MeDIP, CSet = CS_Male_CSF, ISet1 = Female_CSF_Input, ISet2 = Male_CSF_Input, p.adj = "none", diff.method = "edgeR", MeDIP = T, CNV = T, minRowSum = 10)
```


#### Female CSF1 vs Female control

```{r}
Female_csf_vs_Female_control_mr.edgeR.s = MEDIPS.selectSig(results = Female_csf_vs_Female_control_mr.edgeR, p.value = 0.05, adj = F, ratio = 2, bg.counts = 5, CNV = T)
```

#### Male CSF1 vs Male control

```{r}
Male_csf_vs_Male_control_mr.edgeR.s = MEDIPS.selectSig(results = Male_csf_vs_Male_control_mr.edgeR, p.value = 0.05, adj = F, ratio = 2, bg.counts = 5, CNV = T)
```

#### Female control vs Male control

```{r}
Female_control_vs_Male_control_mr.edgeR.s = MEDIPS.selectSig(results = Female_control_vs_Male_control_mr.edgeR, p.value = 0.05, adj = F, ratio = 2, bg.counts = 5, CNV = T)
```

#### Female CSF vs Male CSF

```{r}
Female_csf_vs_Male_csf_mr.edgeR.s = MEDIPS.selectSig(results = Female_csf_vs_Male_csf_mr.edgeR, p.value = 0.05, adj = F, ratio = 2, bg.counts = 5, CNV = T)
```

# 5. Postprocess

```{r, results= 'hide'}
library(magrittr)
library(dplyr)
library(clusterProfiler)
library(org.Rn.eg.db)
```

```{r, results= 'hide'}
source(file.path(SCRIPT_DIR, "rn6_annotation.R"))
rn6_gtf <- input_file(
  "annotations", "rn6", "Rattus_norvegicus.Rnor_6.0.104.gtf.gz"
)
assert_rn6_bam_headers(bam_files, BSgenome.Rnorvegicus.UCSC.rn6)
rn6_annotation <- build_rn6_medips_annotations(rn6_gtf, org.Rn.eg.db)
anno.mart.TSS <- rn6_annotation$annotations$TSS
anno.mart.EXON <- rn6_annotation$annotations$EXON
anno.mart.GENE <- rn6_annotation$annotations$GENE

# Export the rn6 Csf1r locus plus 3 kb of flanking context for the Figure 2d
# CpG-density track. This keeps the figure generator independent of live genome
# services while preserving the exact assembly used by MEDIPS.
csf1r_context_start <- 56414488L - 3000L
csf1r_context_stop <- 56458300L + 3000L
csf1r_context <- Biostrings::DNAStringSet(Biostrings::subseq(
  BSgenome.Rnorvegicus.UCSC.rn6::Rnorvegicus[["chr18"]],
  start = csf1r_context_start,
  end = csf1r_context_stop
))
names(csf1r_context) <- paste0(
  "rn6_chr18:", csf1r_context_start, "-", csf1r_context_stop
)
Biostrings::writeXStringSet(
  csf1r_context,
  filepath = table_file("csf1r_context_rn6.fasta"),
  format = "fasta"
)
annotation_information <- rn6_annotation$information
```

```{r}
annotate_windows <- function(regions, anno, region_type, cols_sel,
                             fc_thresh = 0.5, pval_thresh = 0.05) {
  obj <- MEDIPS.setAnnotation(regions = regions, annotation = anno)
  obj$diffexpressed <- "NO"
  obj$diffexpressed[obj$edgeR.logFC > fc_thresh & obj$edgeR.p.value < pval_thresh] <- "Hyper"
  obj$diffexpressed[obj$edgeR.logFC < -fc_thresh & obj$edgeR.p.value < pval_thresh] <- "Hypo"
  obj$delabel <- NA
  obj$delabel[obj$diffexpressed != "NO"] <- obj$diffexpressed[obj$diffexpressed != "NO"]
  df <- obj[, cols_sel]
  df <- df[!is.na(df$`1_id`), ]
  df$ID <- as.character(df$`1_id`)
  gene_info <- annotation_information[[region_type]]
  gene_info <- gene_info[!duplicated(gene_info$ID), , drop = FALSE]
  df <- df %>%
    left_join(gene_info, by = "ID") %>%
    arrange(ID)
  df
}
```

```{r}
cols_sel_FC_FCSF <- c(
  "chr", "start", "stop", "CF",       
  "Female.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.control.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.CSF.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.control.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.CSF.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "MSets1.counts.mean",
  "MSets1.rpkm.mean",
  "MSets1.rms.mean",
  "MSets2.counts.mean",
  "MSets2.rpkm.mean",
  "MSets2.rms.mean",
  "edgeR.logFC",
  "edgeR.logCPM",
  "edgeR.p.value",
  "edgeR.adj.p.value",
  "CNV.log2.ratio",
  "1_id",
  "diffexpressed",
  "delabel"
  ) 
```

```{r}
cols_sel_MC_MCSF <- c(
  "chr", "start", "stop", "CF",       
  "Male.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.control.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.CSF.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.control.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.CSF.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "MSets1.counts.mean",
  "MSets1.rpkm.mean",
  "MSets1.rms.mean",
  "MSets2.counts.mean",
  "MSets2.rpkm.mean",
  "MSets2.rms.mean",
  "edgeR.logFC",
  "edgeR.logCPM",
  "edgeR.p.value",
  "edgeR.adj.p.value",
  "CNV.log2.ratio",
  "1_id",
  "diffexpressed",
  "delabel"
  ) 
```

```{r}
cols_sel_FC_MC <- c(
  "chr", "start", "stop", "CF",       
  "Female.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.control.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.control.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.control.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.control.input_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "MSets1.counts.mean",
  "MSets1.rpkm.mean",
  "MSets1.rms.mean",
  "MSets2.counts.mean",
  "MSets2.rpkm.mean",
  "MSets2.rms.mean",
  "edgeR.logFC",
  "edgeR.logCPM",
  "edgeR.p.value",
  "edgeR.adj.p.value",
  "CNV.log2.ratio",
  "1_id",
  "diffexpressed",
  "delabel"
  ) 
```

```{r}
cols_sel_FCSF_MCSF <- c(
  "chr", "start", "stop", "CF",       
  "Female.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Male.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.counts",
  "Female.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Male.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rpkm",
  "Female.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Female.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "Male.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms",
  "MSets1.counts.mean",
  "MSets1.rpkm.mean",
  "MSets1.rms.mean",
  "MSets2.counts.mean",
  "MSets2.rpkm.mean",
  "MSets2.rms.mean",
  "edgeR.logFC",
  "edgeR.logCPM",
  "edgeR.p.value",
  "edgeR.adj.p.value",
  "CNV.log2.ratio",
  "1_id",
  "diffexpressed",
  "delabel"
  ) 
```

```{r}
# 1. Female CSF1 vs Female control
F_CSF_vs_F_C_TSS  <- annotate_windows(
  regions   = Female_csf_vs_Female_control_mr.edgeR.s,
  anno      = anno.mart.TSS,
  region_type = "TSS",
  cols_sel  = cols_sel_FC_FCSF
)
F_CSF_vs_F_C_EXON <- annotate_windows(
  regions   = Female_csf_vs_Female_control_mr.edgeR.s,
  anno      = anno.mart.EXON,
  region_type = "EXON",
  cols_sel  = cols_sel_FC_FCSF
)
F_CSF_vs_F_C_GENE <- annotate_windows(
  regions   = Female_csf_vs_Female_control_mr.edgeR.s,
  anno      = anno.mart.GENE,
  region_type = "GENE",
  cols_sel  = cols_sel_FC_FCSF
)

# 2. Male CSF1 vs Male control
M_CSF_vs_M_C_TSS  <- annotate_windows(
  regions   = Male_csf_vs_Male_control_mr.edgeR.s,
  anno      = anno.mart.TSS,
  region_type = "TSS",
  cols_sel  = cols_sel_MC_MCSF
)
M_CSF_vs_M_C_EXON <- annotate_windows(
  regions   = Male_csf_vs_Male_control_mr.edgeR.s,
  anno      = anno.mart.EXON,
  region_type = "EXON",
  cols_sel  = cols_sel_MC_MCSF
)
M_CSF_vs_M_C_GENE <- annotate_windows(
  regions   = Male_csf_vs_Male_control_mr.edgeR.s,
  anno      = anno.mart.GENE,
  region_type = "GENE",
  cols_sel  = cols_sel_MC_MCSF
)

# 3. Female control vs Male control
F_C_vs_M_C_TSS  <- annotate_windows(
  regions   = Female_control_vs_Male_control_mr.edgeR.s,
  anno      = anno.mart.TSS,
  region_type = "TSS",
  cols_sel  = cols_sel_FC_MC
)
F_C_vs_M_C_EXON <- annotate_windows(
  regions   = Female_control_vs_Male_control_mr.edgeR.s,
  anno      = anno.mart.EXON,
  region_type = "EXON",
  cols_sel  = cols_sel_FC_MC
)
F_C_vs_M_C_GENE <- annotate_windows(
  regions   = Female_control_vs_Male_control_mr.edgeR.s,
  anno      = anno.mart.GENE,
  region_type = "GENE",
  cols_sel  = cols_sel_FC_MC
)

# 4. Female CSF vs Male CSF
F_CSF_vs_M_CSF_TSS  <- annotate_windows(
  regions   = Female_csf_vs_Male_csf_mr.edgeR.s,
  anno      = anno.mart.TSS,
  region_type = "TSS",
  cols_sel  = cols_sel_FCSF_MCSF
)
F_CSF_vs_M_CSF_EXON <- annotate_windows(
  regions   = Female_csf_vs_Male_csf_mr.edgeR.s,
  anno      = anno.mart.EXON,
  region_type = "EXON",
  cols_sel  = cols_sel_FCSF_MCSF
)
F_CSF_vs_M_CSF_GENE <- annotate_windows(
  regions   = Female_csf_vs_Male_csf_mr.edgeR.s,
  anno      = anno.mart.GENE,
  region_type = "GENE",
  cols_sel  = cols_sel_FCSF_MCSF
)
```

```{r}
library(dplyr)
library(purrr)
library(stringr)
library(tibble)

df_names <- c(
  "F_CSF_vs_F_C_EXON",  "F_CSF_vs_F_C_TSS",  "F_CSF_vs_F_C_GENE",
  "F_C_vs_M_C_EXON",    "F_C_vs_M_C_TSS",    "F_C_vs_M_C_GENE",
  "M_CSF_vs_M_C_EXON",  "M_CSF_vs_M_C_TSS",  "M_CSF_vs_M_C_GENE",
  "F_CSF_vs_M_CSF_EXON","F_CSF_vs_M_CSF_TSS","F_CSF_vs_M_CSF_GENE"
)
genes_total <- map_dfr(df_names, function(name) {
  df <- as_tibble(get(name))
  
  cols_needed <- c(
    "external_transcript_name", "external_gene_name",
    "transcript_biotype",       "gene_biotype"
  )
  for(col in setdiff(cols_needed, names(df))) {
    df[[col]] <- NA_character_
  }
  
  comparison <- str_remove(name, "_[^_]+$")
  region     <- str_extract(name, "[^_]+$")
  
  df %>%
    dplyr::mutate(
      external_transcript_gene_name = dplyr::coalesce(
        external_transcript_name, external_gene_name
      ),
      transcript_gene_biotype       = dplyr::coalesce(
        transcript_biotype, gene_biotype
      ),
      comparison = comparison,
      region     = region
    ) %>%
    dplyr::select(
      dplyr::everything(),
      -dplyr::any_of(c(
        "external_transcript_name", "external_gene_name",
        "transcript_biotype",       "gene_biotype",
        "ISets2.counts.mean",       "ISets2.rpkm.mean"
      ))
    )
})

GENES_TOTAL <- genes_total %>%
  dplyr::mutate(
    transcript_gene_biotype = factor(transcript_gene_biotype),
    description             = factor(description)
  )
```



```{r}
library(dplyr)
GENES_TOTAL_sig_contig_colap <- GENES_TOTAL %>%
  dplyr::mutate(
    chr   = as.character(chr),
    start = as.integer(start),
    stop  = as.integer(stop)
  ) %>%
  dplyr::filter(
    diffexpressed %in% c("Hyper", "Hypo"),
    !is.na(description),
    edgeR.p.value < 0.05,
    abs(edgeR.logFC) > 1
  ) %>%

  # ---------------------------------------------------------------------------
  # Discard gene-region-comparison groups with inconsistent methylation direction.
  dplyr::group_by(comparison, region, external_transcript_gene_name) %>%
  dplyr::filter(dplyr::n_distinct(diffexpressed) == 1) %>%
  dplyr::ungroup()
```

```{r export-significant-gene-rms-matrix}
sample_metadata <- tibble::tribble(
  ~sample_col, ~group, ~sex, ~condition, ~replicate, ~sample_label,
  "Female.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "FC", "F", "Control", 1L, "FC_1",
  "Female.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "FC", "F", "Control", 2L, "FC_2",
  "Female.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "FC", "F", "Control", 3L, "FC_3",
  "Female.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "FCSF1", "F", "CSF1", 1L, "FCSF1_1",
  "Female.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "FCSF1", "F", "CSF1", 2L, "FCSF1_2",
  "Female.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "FCSF1", "F", "CSF1", 3L, "FCSF1_3",
  "Male.control.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "MC", "M", "Control", 1L, "MC_1",
  "Male.control.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "MC", "M", "Control", 2L, "MC_2",
  "Male.control.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "MC", "M", "Control", 3L, "MC_3",
  "Male.CSF.1.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "MCSF1", "M", "CSF1", 1L, "MCSF1_1",
  "Male.CSF.2.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "MCSF1", "M", "CSF1", 2L, "MCSF1_2",
  "Male.CSF.3.IP_Rattus_norvegicus_RN6_hisat2_alignment_sort.bam.rms", "MCSF1", "M", "CSF1", 3L, "MCSF1_3"
)
if (!all(sample_metadata$sample_col %in% names(GENES_TOTAL))) {
  stop("The annotated MEDIPS table is missing one or more sample RMS columns.")
}

safe_mean <- function(values) {
  if (all(is.na(values))) NA_real_ else mean(values, na.rm = TRUE)
}
significant_genes <- unique(stats::na.omit(
  as.character(GENES_TOTAL_sig_contig_colap$external_transcript_gene_name)
))
window_rms <- GENES_TOTAL %>%
  dplyr::filter(external_transcript_gene_name %in% significant_genes) %>%
  dplyr::group_by(external_transcript_gene_name, chr, start, stop) %>%
  dplyr::summarise(
    dplyr::across(dplyr::all_of(sample_metadata$sample_col), safe_mean),
    .groups = "drop"
  )
significant_gene_rms <- window_rms %>%
  dplyr::group_by(external_transcript_gene_name) %>%
  dplyr::summarise(
    dplyr::across(dplyr::all_of(sample_metadata$sample_col), safe_mean),
    .groups = "drop"
  )
names(significant_gene_rms) <- c("gene", sample_metadata$sample_label)
utils::write.csv(
  significant_gene_rms,
  table_file("significant_gene_rms_matrix_ws400.csv"),
  row.names = FALSE,
  na = ""
)
utils::write.csv(
  sample_metadata,
  table_file("medip_sample_metadata.csv"),
  row.names = FALSE
)
```

## 6. GO enrichment

```{r}
library(dplyr)
library(clusterProfiler)
library(org.Rn.eg.db)

combinations <- GENES_TOTAL_sig_contig_colap %>%
  distinct(comparison, region)

go_results_list <- list()
ontologies <- c("BP", "MF", "CC")
directions <- c("Hyper", "Hypo")

for (i in seq_len(nrow(combinations))) {
  comp <- combinations$comparison[i]
  reg  <- combinations$region[i]

  for (direction in directions) {
    df_base <- GENES_TOTAL_sig_contig_colap %>%
      filter(comparison == comp,
             region     == reg)

    if (direction == "Hyper") {
      df_filt <- df_base %>%
        filter(edgeR.p.value < 0.05,
               edgeR.logFC   >  1,
               transcript_gene_biotype == "protein_coding")
    } else {
      df_filt <- df_base %>%
        filter(edgeR.p.value < 0.05,
               edgeR.logFC   < -1,
               transcript_gene_biotype == "protein_coding")
    }

    if (reg == "GENE") {
      gene_ids_raw <- as.character(df_filt$ID)
      gene_ids <- gene_ids_raw[grepl("^ENSRNOG", gene_ids_raw)]
      keyType  <- "ENSEMBL"
      
    } else if (reg == "TSS") {
      gene_ids_raw <- as.character(df_filt$ID)
      gene_ids     <- gene_ids_raw[grepl("^ENSRNOT", gene_ids_raw)]
      keyType      <- "ENSEMBLTRANS"
      
    } else if (reg == "EXON") {
      gene_ids <- as.character(df_filt$external_transcript_gene_name)
      keyType  <- "SYMBOL"
      
    } else {
      gene_ids <- as.character(df_filt$ID)
      keyType  <- "ENSEMBL"
    }

    gene_ids <- unique(na.omit(gene_ids))
    if (length(gene_ids) == 0) {
      message("No mapped genes for ", comp, " | ", reg, " | ", direction)
      next
    }

    for (ont in ontologies) {
      go_res <- enrichGO(gene          = gene_ids,
                         OrgDb         = org.Rn.eg.db,
                         keyType       = keyType,
                         ont           = ont,
                         readable      = TRUE,
                         pAdjustMethod = "none",
                         qvalueCutoff  = 1,
                         pvalueCutoff  = 0.05)

      df_go <- as.data.frame(go_res)
      if (!is.null(go_res) && nrow(df_go) > 0) {
        key <- paste(comp, reg, direction, ont, sep = "_")
        go_results_list[[key]] <- list(go = go_res)
      } else {
        message("No GO results for ", comp, " | ", reg, " | ", direction, " | ", ont)
      }
    }
  }
}

```

```{r}
# Use qualified dplyr calls to keep data transformations explicit.
library(dplyr)
library(writexl)

write_xlsx(
  GENES_TOTAL_sig_contig_colap,
  table_file("GENES_TOTAL_sig_contig_colap_ws400.xlsx")
)
```

```{r}
combine_go_results_from_list <- function(go_list) {
  out_list <- list()

  for (nm in names(go_list)) {
    parts <- strsplit(nm, "_")[[1]]
    n <- length(parts)

    if (n < 4) {
      warning("The name does not follow the expected pattern: ", nm)
      next
    }

    comparison <- paste(parts[1:(n - 3)], collapse = "_")
    region <- parts[n - 2]
    direction <- parts[n - 1]
    ontology <- parts[n]
    res_df <- go_list[[nm]]$go@result
    res_df$comparison <- comparison
    res_df$region <- region
    res_df$direction <- direction
    res_df$ontology <- ontology
    out_list[[nm]] <- res_df
  }

  bind_rows(out_list)
}

go_results_filtered <- combine_go_results_from_list(go_results_list) %>%
  dplyr::filter(pvalue < 0.05) %>%
  dplyr::mutate(
    p_value = pvalue,
    multiple_testing_adjustment = "none"
  )
if ("geneID" %in% colnames(go_results_filtered)) {
  go_results_filtered$geneID <- gsub("/", ", ", go_results_filtered$geneID)
}
```

```{r go-thematic-summary-and-ranking-ws400}
# GO thematic summary and relevance distribution for 400-bp windows.

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(writexl)
  library(purrr)
  library(tibble)
})

# -----------------------------------------------------------------------------
# 1) Editable dictionary of biological themes
# -----------------------------------------------------------------------------
go_theme_dictionary <- tibble::tribble(
  ~biological_theme, ~pattern,
  "Morphology", "morpho|cell shape|cell projection|process extension|neurite|dendrite|axon|morphogenesis|cytoskelet|actin|microtubule|tubulin|branch",
  "Inflammation / neuroinflammation", "inflam|neuroinflam|immune|microglia|\\bcytokine\\b|interleukin|tnf|nf-kb|complement|toll-like|phagocyt|chemokine",
  "Sex hormones / estrogen", "steroid|estrogen|androgen|hormone|estradiol|progesterone|testosterone|gonadal|\\bsex\\b|aromatase|estrous",
  "Mechanical stimulus / pain", "mechani|mechanosens|\\bpain\\b|nocicepti|touch|somatosensory|sensory stimulus|allodynia"
)

comparison_map <- tibble::tribble(
  ~comparison, ~comparison_label, ~facet_order,
  "F_C_vs_M_C", "GO F vs M", 1L,
  "F_CSF_vs_F_C", "GO F CSF1 vs F C", 2L,
  "M_CSF_vs_M_C", "GO M CSF1 vs M C", 3L
)

direction_levels <- c("Hyper", "Hypo")
theme_levels <- go_theme_dictionary$biological_theme

# -----------------------------------------------------------------------------
# 2) Build master GO table from existing objects
# -----------------------------------------------------------------------------
if (!exists("go_results_filtered")) {
  stop("Object 'go_results_filtered' not found. Run the GO enrichment section first.")
}

go_from_main <- go_results_filtered %>%
  dplyr::transmute(
    comparison = as.character(comparison),
    direction = as.character(direction),
    region = as.character(region),
    Description = as.character(Description),
    p_value = suppressWarnings(as.numeric(pvalue)),
    Count = suppressWarnings(as.numeric(Count)),
    ontology = as.character(ontology),
    geneID = as.character(geneID)
  ) %>%
  dplyr::filter(comparison %in% c("F_C_vs_M_C", "F_CSF_vs_F_C", "M_CSF_vs_M_C"))

go_master_84b <- go_from_main %>%
  dplyr::filter(
    !is.na(Description), Description != "",
    !is.na(direction), direction %in% direction_levels
  ) %>%
  dplyr::left_join(comparison_map, by = "comparison") %>%
  dplyr::filter(!is.na(comparison_label)) %>%
  dplyr::mutate(
    p_value = suppressWarnings(as.numeric(p_value)),
    Count = suppressWarnings(as.numeric(Count)),
    neglog10_pvalue = -log10(pmax(p_value, 1e-300)),
    go_term_key = paste(region, ontology, Description, sep = " | ")
  )

# Diagnostic for missing comparison-direction combinations.
diag_comp_84b <- tidyr::expand_grid(
  comparison = comparison_map$comparison,
  direction = direction_levels
) %>%
  dplyr::left_join(
    go_master_84b %>% dplyr::count(comparison, direction, name = "n_rows"),
    by = c("comparison", "direction")
  ) %>%
  dplyr::mutate(n_rows = dplyr::coalesce(n_rows, 0L)) %>%
  dplyr::arrange(comparison, direction)

print(diag_comp_84b)

# -----------------------------------------------------------------------------
# 3) Theme matching helpers
# -----------------------------------------------------------------------------
match_go_themes_long <- function(df, dict_tbl) {
  purrr::map_dfr(seq_len(nrow(dict_tbl)), function(i) {
    th <- dict_tbl$biological_theme[i]
    patt <- dict_tbl$pattern[i]
    df %>%
      dplyr::filter(stringr::str_detect(Description, stringr::regex(patt, ignore_case = TRUE))) %>%
      dplyr::mutate(biological_theme = th)
  }) %>%
    dplyr::distinct(
      comparison, comparison_label, facet_order, direction, region, ontology,
      Description, p_value, Count, neglog10_pvalue, go_term_key, biological_theme,
      .keep_all = FALSE
    )
}

assign_primary_theme <- function(df, dict_tbl) {
  primary_theme <- purrr::map_chr(df$Description, function(desc_i) {
    matches <- dict_tbl$biological_theme[
      purrr::map_lgl(dict_tbl$pattern, ~ stringr::str_detect(desc_i, stringr::regex(.x, ignore_case = TRUE)))
    ]
    if (length(matches) == 0) "Other" else matches[[1]]
  })
  df %>% dplyr::mutate(biological_theme = primary_theme)
}

go_interest_long_84b <- match_go_themes_long(go_master_84b, go_theme_dictionary)

go_rank_table_84b <- assign_primary_theme(go_master_84b, go_theme_dictionary) %>%
  dplyr::group_by(comparison, comparison_label, facet_order, direction) %>%
  dplyr::arrange(dplyr::desc(neglog10_pvalue), dplyr::desc(Count), Description, .by_group = TRUE) %>%
  dplyr::mutate(rank_within_group = dplyr::row_number()) %>%
  dplyr::ungroup()

# -----------------------------------------------------------------------------
# Ranking/distribution by relevance
# Ranking metric = -log10(unadjusted p value), descending.
# Ranked scatter is chosen over a plain histogram because it preserves location
# of terms of interest across the full relevance order.
# -----------------------------------------------------------------------------
go_rank_export_84b <- go_rank_table_84b %>%
  dplyr::mutate(
    comparison_label = factor(comparison_label, levels = comparison_map$comparison_label),
    direction = factor(direction, levels = direction_levels),
    biological_theme = factor(biological_theme, levels = c(theme_levels, "Other"))
  ) %>%
  dplyr::select(
    comparison,
    comparison_label,
    direction,
    region,
    Description,
    p_value,
    neglog10_pvalue,
    Count,
    biological_theme,
    rank_within_group,
    ontology,
    geneID
  ) %>%
  dplyr::arrange(comparison_label, direction, rank_within_group)

write_xlsx(
  go_rank_export_84b,
  path = table_file("GO_relevance_rank_distribution_ws400.xlsx")
)
message("[OK] GO_relevance_rank_distribution_ws400.xlsx saved")

# The final figure code uses GO_relevance_rank_distribution_ws400.xlsx only.
```

```{r candidate-gene-aliases}
# =============================================================================
# Table 1: manually curated aliases
# For each gene/family of interest, define a canonical name and
# symbols to search in external_transcript_gene_name.
# =============================================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(writexl)
})

gene_aliases <- tibble::tribble(
  ~family,        ~symbol,     ~notes,
  "CSF1R",        "Csf1r",     "Colony stimulating factor 1 receptor",
  "Aromatase",    "Cyp19a1",   "Cytochrome P450 19A1",
  "ERs",          "Esr1",      "Estrogen receptor 1",
  "ERs",          "Esr2",      "Estrogen receptor 2",
  "ERs",          "Gper1",     "G protein-coupled estrogen receptor 1",
  "Dnmts",        "Dnmt1",     "DNA methyltransferase 1",
  "Dnmts",        "Dnmt3a",    "DNA methyltransferase 3A",
  "Dnmts",        "Dnmt3b",    "DNA methyltransferase 3B",
  "Gadds",        "Gadd45a",   "Growth arrest / DNA damage inducible alpha",
  "Gadds",        "Gadd45b",   "Growth arrest / DNA damage inducible beta",
  "Gadds",        "Gadd45g",   "Growth arrest / DNA damage inducible gamma",
  "TNFa",         "Tnf",       "Tumor necrosis factor (nomenclatura rata)",
  "IL10",         "Il10",      "Interleukin 10",
  "Bdnf",         "Bdnf",      "Brain-derived neurotrophic factor",
  "IRF",          "Irf1",      "Interferon regulatory factor 1",
  "IRF",          "Irf2",      "Interferon regulatory factor 2",
  "IRF",          "Irf3",      "Interferon regulatory factor 3",
  "IRF",          "Irf4",      "Interferon regulatory factor 4",
  "IRF",          "Irf5",      "Interferon regulatory factor 5",
  "IRF",          "Irf6",      "Interferon regulatory factor 6",
  "IRF",          "Irf7",      "Interferon regulatory factor 7",
  "IRF",          "Irf8",      "Interferon regulatory factor 8",
  "IRF",          "Irf9",      "Interferon regulatory factor 9",
  "PU.1",         "Spi1",      "SPI1 / PU.1 transcription factor"
)

# Display alias table
message("=== Alias table (", nrow(gene_aliases), " symbols across ",
        n_distinct(gene_aliases$family), " families) ===")
print(gene_aliases)

```


```{r candidate-gene-search}
# =============================================================================
# Table 2: dataset matches
# Search each symbol in GENES_TOTAL (primary source: 400-bp windows with
# genomic coordinates, including non-significant windows).
# GENES_TOTAL_sig_contig_colap is used as the secondary summary.
# =============================================================================

symbols_to_search <- tolower(gene_aliases$symbol)

# --- Search GENES_TOTAL (primary, higher-resolution source) ---
gene_hits_full <- GENES_TOTAL %>%
  dplyr::filter(
    tolower(external_transcript_gene_name) %in% symbols_to_search
  ) %>%
  dplyr::left_join(
    gene_aliases %>% dplyr::mutate(symbol_lower = tolower(symbol)),
    by = c("external_transcript_gene_name" = "symbol")
  )

# --- Search GENES_TOTAL_sig_contig_colap (significant rows only) ---
gene_hits_sig <- GENES_TOTAL_sig_contig_colap %>%
  dplyr::filter(
    tolower(external_transcript_gene_name) %in% symbols_to_search
  ) %>%
  dplyr::left_join(
    gene_aliases %>% dplyr::mutate(symbol_lower = tolower(symbol)),
    by = c("external_transcript_gene_name" = "symbol")
  )

# --- Summary of detected and undetected symbols ---
found_symbols <- unique(tolower(gene_hits_full$external_transcript_gene_name))
found_sig_symbols <- unique(tolower(gene_hits_sig$external_transcript_gene_name))

status_table <- gene_aliases %>%
  dplyr::mutate(
    found_in_GENES_TOTAL = tolower(symbol) %in% found_symbols,
    found_significant = tolower(symbol) %in% found_sig_symbols,
    n_windows_total = purrr::map_int(tolower(symbol), function(s) {
      sum(tolower(GENES_TOTAL$external_transcript_gene_name) == s, na.rm = TRUE)
    }),
    n_windows_sig = purrr::map_int(tolower(symbol), function(s) {
      sum(tolower(GENES_TOTAL_sig_contig_colap$external_transcript_gene_name) == s, na.rm = TRUE)
    })
  )

message("\n=== Candidate-gene search status ===")
message("Found in GENES_TOTAL: ",
        sum(status_table$found_in_GENES_TOTAL), "/", nrow(status_table))
message("With significant changes: ",
        sum(status_table$found_significant), "/", nrow(status_table))
not_found <- status_table %>% dplyr::filter(!found_in_GENES_TOTAL)
if (nrow(not_found) > 0) {
  message("\nGenes not found in the dataset:")
  for (i in seq_len(nrow(not_found))) {
    message("  - ", not_found$family[i], " (", not_found$symbol[i], ")")
  }
}

print(status_table %>% dplyr::select(family, symbol, found_in_GENES_TOTAL,
                                      found_significant, n_windows_total, n_windows_sig))
```


```{r candidate-gene-hits-export}
# =============================================================================
# Export window-level match table
# =============================================================================

if (nrow(gene_hits_full) > 0) {
  hits_export <- gene_hits_full %>%
    dplyr::select(
      family, symbol = external_transcript_gene_name,
      comparison, region, chr, start, stop,
      edgeR.logFC, edgeR.p.value, edgeR.logCPM,
      CNV.log2.ratio, diffexpressed
    ) %>%
    dplyr::arrange(family, symbol, comparison, region, chr, start)

  write_xlsx(
    hits_export,
    path = table_file("methylation_map_candidate_genes_hits_ws400.xlsx")
  )
  message("[OK] methylation_map_candidate_genes_hits_ws400.xlsx: ",
          nrow(hits_export), " windows")

  # Reproduce the row and column orders used by the final Python heatmap.
  heatmap_gene_comp <- GENES_TOTAL_sig_contig_colap %>%
    dplyr::mutate(
      gene = as.character(external_transcript_gene_name),
      gene_lower = tolower(gene),
      comparison = as.character(comparison),
      edgeR.logFC = suppressWarnings(as.numeric(edgeR.logFC)),
      diffexpressed = as.character(diffexpressed)
    ) %>%
    dplyr::filter(
      diffexpressed %in% c("Hyper", "Hypo"),
      gene_lower %in% tolower(gene_aliases$symbol),
      !is.na(gene), gene != "", !is.na(comparison), !is.na(edgeR.logFC)
    ) %>%
    dplyr::group_by(gene, comparison) %>%
    dplyr::summarise(gene_logFC = mean(edgeR.logFC), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = comparison, values_from = gene_logFC, values_fill = 0)

  heatmap_matrix <- as.matrix(
    heatmap_gene_comp[, setdiff(names(heatmap_gene_comp), "gene"), drop = FALSE]
  )
  rownames(heatmap_matrix) <- heatmap_gene_comp$gene
  variable_rows <- apply(heatmap_matrix, 1, stats::var) > 0
  heatmap_matrix <- heatmap_matrix[variable_rows, , drop = FALSE]

  correlation_order <- function(matrix_input, rows = TRUE) {
    if ((rows && nrow(matrix_input) < 2) || (!rows && ncol(matrix_input) < 2)) {
      return(if (rows) seq_len(nrow(matrix_input)) else seq_len(ncol(matrix_input)))
    }
    correlation_matrix <- if (rows) {
      suppressWarnings(stats::cor(t(matrix_input), use = "pairwise.complete.obs"))
    } else {
      suppressWarnings(stats::cor(matrix_input, use = "pairwise.complete.obs"))
    }
    correlation_matrix[!is.finite(correlation_matrix)] <- 0
    diag(correlation_matrix) <- 1
    stats::hclust(stats::as.dist(1 - correlation_matrix), method = "average")$order
  }

  scaled_heatmap_matrix <- t(scale(t(heatmap_matrix)))
  scaled_heatmap_matrix[!is.finite(scaled_heatmap_matrix)] <- 0
  row_order <- correlation_order(scaled_heatmap_matrix, rows = TRUE)
  column_order <- correlation_order(scaled_heatmap_matrix, rows = FALSE)

  write_xlsx(
    list(
      row_order = tibble::tibble(
        row_id = rownames(scaled_heatmap_matrix)[row_order],
        order = seq_along(row_order)
      ),
      column_order = tibble::tibble(
        column_id = colnames(scaled_heatmap_matrix)[column_order],
        order = seq_along(column_order)
      )
    ),
    path = table_file("heatmap_candidate_genes_logFC_ws400_orders.xlsx")
  )
  message("[OK] heatmap_candidate_genes_logFC_ws400_orders.xlsx saved")
} else {
  message("[!] No candidate-gene windows were found in GENES_TOTAL")
}
```

```{r ere-motif-scan-ws400, fig.width=14, fig.height=9}
# =============================================================================
# ERE motif scan
# Exports motif matches, candidate windows and scan diagnostics.
# =============================================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(purrr)
  library(tibble)
  library(writexl)
  library(GenomicRanges)
})

# Excel cannot store cells longer than 32,767 characters. GO tables can
# accumulate long gene/term lists; to retain traceability this workflow
# saves a complete RDS and a readable XLSX with truncated cells.
# -----------------------------------------------------------------------------
# A) ERE motif scan in candidate-gene DMRs
# -----------------------------------------------------------------------------

ere_candidate_genes <- tibble::tribble(
  ~family,        ~symbol,   ~notes,
  "CSF1R",        "Csf1r",   "Colony stimulating factor 1 receptor",
  "Aromatase",    "Cyp19a1", "Cytochrome P450 19A1 / aromatase",
  "ERs",          "Esr1",    "Estrogen receptor 1",
  "ERs",          "Esr2",    "Estrogen receptor 2",
  "ERs",          "Gper1",   "G protein-coupled estrogen receptor 1",
  "Dnmts",        "Dnmt1",   "DNA methyltransferase 1",
  "Dnmts",        "Dnmt3a",  "DNA methyltransferase 3A",
  "Dnmts",        "Dnmt3b",  "DNA methyltransferase 3B",
  "Gadds",        "Gadd45a", "Growth arrest / DNA damage inducible alpha",
  "Gadds",        "Gadd45b", "Growth arrest / DNA damage inducible beta",
  "Gadds",        "Gadd45g", "Growth arrest / DNA damage inducible gamma",
  "TNFa",         "Tnf",     "Tumor necrosis factor",
  "IL10",         "Il10",    "Interleukin 10",
  "Bdnf",         "Bdnf",    "Brain-derived neurotrophic factor",
  "IRF",          "Irf1",    "Interferon regulatory factor 1",
  "IRF",          "Irf2",    "Interferon regulatory factor 2",
  "IRF",          "Irf3",    "Interferon regulatory factor 3",
  "IRF",          "Irf4",    "Interferon regulatory factor 4",
  "IRF",          "Irf5",    "Interferon regulatory factor 5",
  "IRF",          "Irf6",    "Interferon regulatory factor 6",
  "IRF",          "Irf7",    "Interferon regulatory factor 7",
  "IRF",          "Irf8",    "Interferon regulatory factor 8",
  "IRF",          "Irf9",    "Interferon regulatory factor 9",
  "PU.1",         "Spi1",    "SPI1 / PU.1 transcription factor"
) %>%
  dplyr::mutate(symbol_lower = tolower(symbol))

# ERE motif definitions used in the exact genomic-sequence scan.
# canonical_full_15bp: conservative full ERE, AGGTCA + 3 nt + TGACCT.

ere_motif_definitions <- tibble::tribble(
  ~motif_name,              ~motif_class,                         ~regex,                    ~strand_label, ~interpretation,
  "canonical_full_15bp",    "canonical_full_ERE",                 "AGGTCA[ACGT]{3}TGACCT",   "both",        "Exact full ERE palindrome; conservative",
  "canonical_core_13bp",    "canonical_core_ERE",                 "GGTCA[ACGT]{3}TGACC",     "both",        "Exact 13 bp core ERE; conservative but slightly shorter",
  "half_site_AGGTCA",       "half_site_exact_exploratory",        "AGGTCA",                  "+",           "Exact ER half-site; exploratory",
  "half_site_TGACCT",       "half_site_exact_exploratory",        "TGACCT",                  "-",           "Exact reverse-complement half-site; exploratory"
)

region_to_biological_label <- function(x) {
  dplyr::case_when(
    x == "GENE" ~ "GENEBODY",
    TRUE ~ as.character(x)
  )
}

select_ere_dmr_source <- function() {
  candidates <- c("GENES_TOTAL_sig_contig_colap", "GENES_TOTAL")
  for (nm in candidates) {
    if (!exists(nm, inherits = TRUE)) next
    obj <- get(nm, inherits = TRUE)
    if (!is.data.frame(obj)) next
    required <- c(
      "comparison", "region", "external_transcript_gene_name",
      "diffexpressed", "chr", "start", "stop", "edgeR.logFC",
      "edgeR.p.value"
    )
    if (all(required %in% names(obj))) {
      return(list(source_name = nm, data = tibble::as_tibble(obj)))
    }
  }
  list(source_name = NA_character_, data = tibble::tibble())
}

get_rn6_genome_or_null <- function() {
  if (!requireNamespace("BSgenome.Rnorvegicus.UCSC.rn6", quietly = TRUE)) {
    return(NULL)
  }
  suppressPackageStartupMessages(library(BSgenome.Rnorvegicus.UCSC.rn6))
  if (exists("Rnorvegicus", inherits = TRUE)) {
    return(get("Rnorvegicus", inherits = TRUE))
  }
  NULL
}

normalise_seqname <- function(chr, available_seqlevels) {
  chr <- as.character(chr)
  if (is.na(chr) || chr == "") return(NA_character_)
  if (chr %in% available_seqlevels) return(chr)
  chr_prefixed <- paste0("chr", chr)
  if (chr_prefixed %in% available_seqlevels) return(chr_prefixed)
  if (chr %in% c("M", "MT", "chrMT") && "chrM" %in% available_seqlevels) return("chrM")
  NA_character_
}

extract_sequences_vectorized <- function(chr, start, stop, genome_obj) {
  n <- length(chr)
  out <- rep(NA_character_, n)
  if (is.null(genome_obj) || n == 0) return(out)

  available <- seqnames(genome_obj)
  chr_ok <- vapply(chr, normalise_seqname, character(1), available_seqlevels = available)
  start <- suppressWarnings(as.integer(start))
  stop <- suppressWarnings(as.integer(stop))

  valid <- !is.na(chr_ok) & !is.na(start) & !is.na(stop) & start <= stop
  if (!any(valid)) return(out)

  idx <- which(valid)
  seqlen <- seqlengths(genome_obj)[chr_ok[idx]]
  starts <- pmax(1L, start[idx])
  stops <- pmin(stop[idx], as.integer(seqlen))
  keep <- starts <= stops
  if (!any(keep)) return(out)

  idx <- idx[keep]
  gr <- GenomicRanges::GRanges(
    seqnames = chr_ok[idx],
    ranges = IRanges::IRanges(start = starts[keep], end = stops[keep])
  )

  out[idx] <- tryCatch(
    as.character(BSgenome::getSeq(genome_obj, gr)),
    error = function(e) {
      message("[DIAGNOSTIC] Vectorized getSeq failed: ", e$message)
      rep(NA_character_, length(idx))
    }
  )

  out
}

locate_regex_matches <- function(sequence, motif_tbl = ere_motif_definitions) {
  if (is.na(sequence) || sequence == "") {
    return(tibble::tibble())
  }
  sequence <- toupper(sequence)

  purrr::map_dfr(seq_len(nrow(motif_tbl)), function(i) {
    patt <- motif_tbl$regex[i]
    hits <- gregexpr(patt, sequence, perl = TRUE, ignore.case = TRUE)[[1]]
    if (length(hits) == 1 && hits[1] == -1) return(tibble::tibble())

    match_len <- attr(hits, "match.length")
    tibble::tibble(
      motif_name = motif_tbl$motif_name[i],
      motif_class = motif_tbl$motif_class[i],
      motif_regex = patt,
      motif_strand = motif_tbl$strand_label[i],
      motif_start_in_window = as.integer(hits),
      motif_end_in_window = as.integer(hits + match_len - 1L),
      motif_sequence = substr(sequence, hits, hits + match_len - 1L)
    )
  })
}

ere_source <- select_ere_dmr_source()
ere_genome <- get_rn6_genome_or_null()

ere_scan_diagnostic <- tibble::tibble(
  item = c(
    "source_object",
    "genome_available",
    "motif_strategy",
    "canonical_full_definition",
    "canonical_core_definition",
    "half_site_definition",
    "ERE_folder"
  ),
  value = c(
    as.character(ere_source$source_name),
    as.character(!is.null(ere_genome)),
    "Exact regex scan on genomic windows; no mismatch-tolerant PWM threshold used",
    "AGGTCA[ACGT]{3}TGACCT",
    "GGTCA[ACGT]{3}TGACC",
    "Exact AGGTCA / TGACCT only; exploratory, not interpreted as sufficient evidence",
    normalizePath(input_file("ERE"), winslash = "/", mustWork = FALSE)
  )
)

if (nrow(ere_source$data) == 0) {
  ere_scan_table <- tibble::tibble(
    diagnostic = "No suitable DMR/window object with comparison, region, gene symbol and coordinates was found."
  )
  writexl::write_xlsx(
    list(
      ERE_scan = ere_scan_table,
      diagnostic = ere_scan_diagnostic,
      motif_definitions = ere_motif_definitions,
      candidate_genes = ere_candidate_genes
    ),
    path = table_file("DMR_candidate_genes_ERE_scan_ws400.xlsx")
  )
  message("[DIAGNOSTIC] ERE scan not run: no source object with coordinates is available.")
} else {
  dmr_candidate_windows <- ere_source$data %>%
    dplyr::mutate(
      ere_row_id = dplyr::row_number(),
      gene_symbol = as.character(external_transcript_gene_name),
      symbol_lower = tolower(gene_symbol),
      comparison = as.character(comparison),
      region = as.character(region),
      region_biological = region_to_biological_label(region),
      diffexpressed = as.character(diffexpressed),
      methylation_direction = dplyr::case_when(
        diffexpressed == "Hyper" ~ "HYPER",
        diffexpressed == "Hypo" ~ "HYPO",
        TRUE ~ as.character(diffexpressed)
      ),
      chr = as.character(chr),
      start = suppressWarnings(as.integer(start)),
      stop = suppressWarnings(as.integer(stop))
    ) %>%
    dplyr::filter(diffexpressed %in% c("Hyper", "Hypo")) %>%
    dplyr::inner_join(
      ere_candidate_genes %>% dplyr::select(family, symbol, symbol_lower, notes),
      by = "symbol_lower"
    ) %>%
    dplyr::arrange(family, symbol, comparison, region, chr, start) %>%
    dplyr::distinct(
      family, symbol, comparison, region, chr, start, stop,
      edgeR.logFC, edgeR.p.value, .keep_all = TRUE
    ) %>%
    dplyr::mutate(ere_row_id = dplyr::row_number())

  if (nrow(dmr_candidate_windows) == 0) {
    ere_scan_table <- tibble::tibble(
      diagnostic = "No significant DMR/window rows were found for the requested candidate genes.",
      source_object = ere_source$source_name
    )
    writexl::write_xlsx(
      list(
        ERE_scan = ere_scan_table,
        diagnostic = ere_scan_diagnostic,
        motif_definitions = ere_motif_definitions,
        candidate_genes = ere_candidate_genes
      ),
      path = table_file("DMR_candidate_genes_ERE_scan_ws400.xlsx")
    )
    message("[DIAGNOSTIC] ERE scan found no candidate-gene matches in significant DMRs.")
  } else {
    dmr_candidate_windows <- dmr_candidate_windows %>%
      dplyr::mutate(
        coordinate_status = dplyr::case_when(
          is.na(chr) | is.na(start) | is.na(stop) ~ "missing_coordinates",
          is.null(ere_genome) ~ "BSgenome_Rnorvegicus_rn6_not_available",
          TRUE ~ "coordinates_available"
        )
      )

    dmr_candidate_windows$sequence_full <- extract_sequences_vectorized(
      dmr_candidate_windows$chr,
      dmr_candidate_windows$start,
      dmr_candidate_windows$stop,
      ere_genome
    )

    csf1r_windows <- dmr_candidate_windows %>%
      dplyr::filter(
        tolower(symbol) == "csf1r",
        !is.na(sequence_full),
        sequence_full != ""
      ) %>%
      dplyr::distinct(chr, start, stop, sequence_full)
    if (nrow(csf1r_windows) == 0) {
      stop("No rn6 DMR windows overlap the rn6 Csf1r annotation.")
    }
    fasta_lines <- unlist(
      Map(
        function(chr, start, stop, sequence) {
          c(paste0(">", chr, ":", start, "-", stop), sequence)
        },
        csf1r_windows$chr,
        csf1r_windows$start,
        csf1r_windows$stop,
        csf1r_windows$sequence_full
      ),
      use.names = FALSE
    )
    writeLines(fasta_lines, table_file("csf1r_dmr_windows_rn6.fasta"))

    motif_hits_long <- purrr::map2_dfr(
      dmr_candidate_windows$ere_row_id,
      dmr_candidate_windows$sequence_full,
      function(row_id, seq_i) {
        locate_regex_matches(seq_i) %>%
          dplyr::mutate(ere_row_id = row_id)
      }
    )

    if (nrow(motif_hits_long) == 0) {
      motif_hits_long <- tibble::tibble(
        ere_row_id = integer(),
        motif_name = character(),
        motif_class = character(),
        motif_regex = character(),
        motif_strand = character(),
        motif_start_in_window = integer(),
        motif_end_in_window = integer(),
        motif_sequence = character(),
        chr = character(),
        motif_genomic_start = integer(),
        motif_genomic_stop = integer()
      )

      motif_summary <- tibble::tibble(
        ere_row_id = integer(),
        has_ERE_full_15bp = logical(),
        has_ERE_core_13bp = logical(),
        has_ERE_canonical = logical(),
        has_ERE_half_site_exact = logical(),
        n_ERE_full_15bp = integer(),
        n_ERE_core_13bp = integer(),
        n_ERE_half_site_exact = integer(),
        motif_positions_in_window = character(),
        motif_positions_genomic = character(),
        motif_sequences = character()
      )
    } else {
      motif_hits_long <- motif_hits_long %>%
        dplyr::left_join(
          dmr_candidate_windows %>% dplyr::select(ere_row_id, chr, window_start = start),
          by = "ere_row_id"
        ) %>%
        dplyr::mutate(
          motif_genomic_start = window_start + motif_start_in_window - 1L,
          motif_genomic_stop = window_start + motif_end_in_window - 1L
        ) %>%
        dplyr::select(-window_start)

      motif_summary <- motif_hits_long %>%
        dplyr::group_by(ere_row_id) %>%
        dplyr::summarise(
          has_ERE_full_15bp = any(motif_name == "canonical_full_15bp"),
          has_ERE_core_13bp = any(motif_name == "canonical_core_13bp"),
          has_ERE_canonical = any(motif_class %in% c("canonical_full_ERE", "canonical_core_ERE")),
          has_ERE_half_site_exact = any(motif_class == "half_site_exact_exploratory"),
          n_ERE_full_15bp = sum(motif_name == "canonical_full_15bp"),
          n_ERE_core_13bp = sum(motif_name == "canonical_core_13bp"),
          n_ERE_half_site_exact = sum(motif_class == "half_site_exact_exploratory"),
          motif_positions_in_window = paste(
            paste0(motif_name, ":", motif_start_in_window, "-", motif_end_in_window, "(", motif_strand, ")"),
            collapse = "; "
          ),
          motif_positions_genomic = paste(
            paste0(chr, ":", motif_genomic_start, "-", motif_genomic_stop, "(", motif_strand, ")"),
            collapse = "; "
          ),
          motif_sequences = paste(unique(motif_sequence), collapse = "; "),
          .groups = "drop"
        )
    }

    ere_scan_table <- dmr_candidate_windows %>%
      dplyr::left_join(motif_summary, by = "ere_row_id") %>%
      dplyr::mutate(
        has_ERE_full_15bp = dplyr::coalesce(has_ERE_full_15bp, FALSE),
        has_ERE_core_13bp = dplyr::coalesce(has_ERE_core_13bp, FALSE),
        has_ERE_canonical = dplyr::coalesce(has_ERE_canonical, FALSE),
        has_ERE_half_site_exact = dplyr::coalesce(has_ERE_half_site_exact, FALSE),
        n_ERE_full_15bp = dplyr::coalesce(n_ERE_full_15bp, 0L),
        n_ERE_core_13bp = dplyr::coalesce(n_ERE_core_13bp, 0L),
        n_ERE_half_site_exact = dplyr::coalesce(n_ERE_half_site_exact, 0L),
        sequence_status = dplyr::case_when(
          is.na(sequence_full) ~ "sequence_not_available",
          TRUE ~ "sequence_extracted"
        ),
        sequence_analysed = dplyr::case_when(
          is.na(sequence_full) ~ NA_character_,
          nchar(sequence_full) <= 1000 ~ sequence_full,
          TRUE ~ paste0(substr(sequence_full, 1, 500), "...[truncated_length=", nchar(sequence_full), "]")
        ),
        source_object = ere_source$source_name
      ) %>%
      dplyr::select(
        comparison,
        methylation_direction,
        gene = symbol,
        family,
        region = region_biological,
        chr,
        start,
        stop,
        edgeR.logFC,
        edgeR.p.value,
        edgeR.logCPM,
        CNV.log2.ratio,
        sequence_status,
        coordinate_status,
        sequence_analysed,
        has_ERE_full_15bp,
        has_ERE_core_13bp,
        has_ERE_canonical,
        has_ERE_half_site_exact,
        n_ERE_full_15bp,
        n_ERE_core_13bp,
        n_ERE_half_site_exact,
        motif_positions_in_window,
        motif_positions_genomic,
        motif_sequences,
        source_object
      )

    ere_missing_diag <- ere_scan_table %>%
      dplyr::count(sequence_status, coordinate_status, name = "n_windows") %>%
      dplyr::arrange(sequence_status, coordinate_status)

    # Classify exact motif matches separately from other candidate DMR windows.
    ere_window_summary_by_gene <- ere_scan_table %>%
      dplyr::mutate(
        ERE_status = dplyr::case_when(
          has_ERE_canonical ~ "Canonical/core ERE",
          !has_ERE_canonical & has_ERE_half_site_exact ~ "Half-site only",
          TRUE ~ "No exact ERE"
        ),
        ERE_status = factor(
          ERE_status,
          levels = c("No exact ERE", "Half-site only", "Canonical/core ERE")
        )
      ) %>%
      dplyr::count(comparison, family, gene, ERE_status, name = "n_windows") %>%
      dplyr::group_by(comparison, family, gene) %>%
      tidyr::complete(
        ERE_status = factor(
          c("No exact ERE", "Half-site only", "Canonical/core ERE"),
          levels = c("No exact ERE", "Half-site only", "Canonical/core ERE")
        ),
        fill = list(n_windows = 0L)
      ) %>%
      dplyr::ungroup() %>%
      tidyr::pivot_wider(
        names_from = ERE_status,
        values_from = n_windows,
        values_fill = 0
      ) %>%
      dplyr::mutate(
        n_candidate_DMR_windows = `Canonical/core ERE` + `Half-site only` + `No exact ERE`,
        n_windows_with_exact_motif = `Canonical/core ERE` + `Half-site only`
      ) %>%
      dplyr::arrange(comparison, family, gene)

    ere_candidate_gene_status <- ere_candidate_genes %>%
      dplyr::left_join(
        ere_scan_table %>%
          dplyr::group_by(family, gene) %>%
          dplyr::summarise(
            n_candidate_DMR_windows = dplyr::n(),
            n_windows_with_canonical_or_core = sum(has_ERE_canonical, na.rm = TRUE),
            n_windows_with_half_site_only = sum(!has_ERE_canonical & has_ERE_half_site_exact, na.rm = TRUE),
            n_windows_no_exact_motif = sum(!has_ERE_canonical & !has_ERE_half_site_exact, na.rm = TRUE),
            comparisons_with_candidate_DMR = paste(sort(unique(comparison)), collapse = "; "),
            .groups = "drop"
          ) %>%
          dplyr::rename(symbol = gene),
        by = c("family", "symbol")
      ) %>%
      dplyr::mutate(
        dplyr::across(
          c(
            n_candidate_DMR_windows,
            n_windows_with_canonical_or_core,
            n_windows_with_half_site_only,
            n_windows_no_exact_motif
          ),
          ~ dplyr::coalesce(.x, 0L)
        ),
        ERE_scan_status = dplyr::case_when(
          n_candidate_DMR_windows == 0 ~ "no_significant_candidate_DMR_window",
          n_windows_with_canonical_or_core > 0 ~ "candidate_DMR_with_canonical_or_core_ERE",
          n_windows_with_half_site_only > 0 ~ "candidate_DMR_with_half_site_only",
          TRUE ~ "candidate_DMR_without_exact_ERE_motif"
        )
      ) %>%
      dplyr::arrange(family, symbol)

    writexl::write_xlsx(
      list(
        ERE_scan = ere_scan_table,
        motif_hits_long = motif_hits_long,
        ERE_window_summary_by_gene = ere_window_summary_by_gene,
        ERE_candidate_gene_status = ere_candidate_gene_status,
        diagnostic = ere_scan_diagnostic,
        missing_sequence_summary = ere_missing_diag,
        motif_definitions = ere_motif_definitions,
        candidate_genes = ere_candidate_genes
      ),
      path = table_file("DMR_candidate_genes_ERE_scan_ws400.xlsx")
    )
    message("[OK] DMR_candidate_genes_ERE_scan_ws400.xlsx saved")

  }
}
```

```{r validate-release-tables}
required_release_tables <- c(
  "GENES_TOTAL_sig_contig_colap_ws400.xlsx",
  "GO_relevance_rank_distribution_ws400.xlsx",
  "methylation_map_candidate_genes_hits_ws400.xlsx",
  "heatmap_candidate_genes_logFC_ws400_orders.xlsx",
  "DMR_candidate_genes_ERE_scan_ws400.xlsx",
  "significant_gene_rms_matrix_ws400.csv",
  "medip_sample_metadata.csv",
  "csf1r_dmr_windows_rn6.fasta",
  "csf1r_context_rn6.fasta"
)

required_release_paths <- table_file(required_release_tables)
missing_release_tables <- required_release_paths[!file.exists(required_release_paths)]
if (length(missing_release_tables) > 0) {
  stop(
    "The MEDIPS run did not generate the following required release tables:\n",
    paste(missing_release_tables, collapse = "\n")
  )
}

release_info <- file.info(required_release_paths)
stale_release_tables <- rownames(release_info)[
  is.na(release_info$mtime) |
    release_info$mtime < (RUN_STARTED_AT - 2) |
    release_info$size <= 0
]
if (length(stale_release_tables) > 0) {
  stop(
    "The following release tables were not refreshed during this run:\n",
    paste(stale_release_tables, collapse = "\n")
  )
}

required_sheets <- list(
  "DMR_candidate_genes_ERE_scan_ws400.xlsx" = c("ERE_scan", "candidate_genes"),
  "heatmap_candidate_genes_logFC_ws400_orders.xlsx" = c("row_order", "column_order")
)
for (workbook_name in names(required_sheets)) {
  workbook_path <- table_file(workbook_name)
  missing_sheets <- setdiff(required_sheets[[workbook_name]], readxl::excel_sheets(workbook_path))
  if (length(missing_sheets) > 0) {
    stop(
      "Workbook ", workbook_name, " is missing required sheets: ",
      paste(missing_sheets, collapse = ", ")
    )
  }
}

message(
  "[OK] All MEDIPS release tables were regenerated in: ",
  normalizePath(TABLE_DIR, winslash = "/", mustWork = TRUE)
)
```


